---
title: "scQCenrich QC Report"
subtitle: "Automated metrics, flags and diagnostics"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
    number_sections: false
    self_contained: true
---



```{r report-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = NA     # ← removes the "## " prefix on outputs
)

# Objects provided by render_qc_report() via environment:
# - obj: Seurat object
# - metrics: data.frame of QC metrics
# - status_df: data.frame of flags (qc_status / etc.)
# - save_dir: directory containing pre-rendered PNGs/CSVs
# - SCOL: optional sample column name (character or NULL)
# - theme_bootswatch: bootswatch theme name (character)

suppressPackageStartupMessages({
  library(ggplot2)
  BG <- getOption("scQCenrich.plot_bg", "white")  # "white" or "transparent"

  if (requireNamespace("bslib", quietly = TRUE)) {
    theme <- bslib::bs_theme(
      bootswatch   = get0("theme_bootswatch", ifnotfound = "minty"),
      base_font    = bslib::font_google("Inter", local = TRUE),
      heading_font = bslib::font_google("Inter", local = TRUE)
    )
    options(htmltools.theme = theme)
  }
  if (requireNamespace("thematic", quietly = TRUE)) {
    # lock background color for all figures produced during Rmd render
    thematic::thematic_rmd(bg = BG, fg = "black")
  }
  # ggplot default look with white panel & plot background
  theme_set(theme_classic(base_size = 11) +
              theme(plot.background  = element_rect(fill = BG, colour = NA),
                    panel.background = element_rect(fill = BG, colour = NA)))
})

```

> **Overview**

* Date: `r format(Sys.time(), "%Y-%m-%d %H:%M")`
* Genes: `r nrow(Seurat::GetAssayData(obj, layer = "counts"))`
* Cells: `r ncol(Seurat::GetAssayData(obj, layer = "counts"))`
* Kept cells: `r sum(status_df$qc_status == "keep")`
* Borderline cells: `r sum(status_df$qc_status == "borderline")`
* Removed cells: `r sum(status_df$qc_status == "remove")`

```{r badges}
library(htmltools)
kept <- sum(status_df$qc_status == "keep")
dropped <- sum(status_df$qc_status == "remove")
borderline  <- sum(status_df$qc_status == "borderline")

dbls <- sum(as.logical(obj@meta.data$is_doublet %||% FALSE), na.rm = TRUE)
div(class = "d-flex gap-2",
  tags$span(class="badge text-bg-success", paste("KEEP", kept)),
  tags$span(class="badge text-bg-danger",  paste("REMOVE", dropped)),
    tags$span(class="badge text-bg-warning",  paste("BORDERLINE", borderline)),

  tags$span(class="badge text-bg-warning", paste("DOUBLETS", dbls))
)
```

### UMAP (colored by QC status)

```{r umap-qc, fig.width=7, fig.height=6}
umap_png <- file.path(save_dir, "qc_umap_qc.png")
if (file.exists(umap_png)) knitr::include_graphics(umap_png) else {
  plot.new(); text(0.5, 0.5, "UMAP image not found; run qc_diagnostic_panel() first.")
}
```


### Clustering (used for annotation)

```{r umap-clusters, fig.width=7, fig.height=6}
BG <- getOption("scQCenrich.plot_bg", "white")

# Reuse/create UMAP & clusters as before...
if (!("umap" %in% names(obj@reductions))) {
  obj <- scQCenrich:::.ensure_visual_embeddings(obj, pc_dims = 1:20)
}
if (!("seurat_clusters" %in% colnames(obj@meta.data))) {
  obj <- Seurat::FindNeighbors(obj, dims = 1:20, verbose = FALSE)
  obj <- Seurat::FindClusters(obj, resolution = 0.4, verbose = FALSE)
}

# Try a normal labeled plot first (may fail on some ggplot2/Seurat combos)
p_try <- try(
  Seurat::DimPlot(obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE) +
    ggplot2::ggtitle("UMAP (Seurat clusters)") +
    ggplot2::theme_classic(base_size = 11) +
    ggplot2::theme(
      plot.background  = ggplot2::element_rect(fill = BG, colour = NA),
      panel.background = ggplot2::element_rect(fill = BG, colour = NA)
    ),
  silent = TRUE
)


if (!inherits(p_try, "try-error")) {
  p_try
} else {
  # Robust manual labeling
  emb <- Seurat::Embeddings(obj, "umap")
  cl  <- as.character(obj$seurat_clusters)
  ok  <- is.finite(emb[, 1]) & is.finite(emb[, 2]) & !is.na(cl)
  emb <- emb[ok, , drop = FALSE]
  cl  <- cl[ok]

  # rowsum-based centers (works even if colnames are odd)
  if (nrow(emb) && length(cl)) {
    n_by <- as.vector(table(cl)[rownames(rowsum(emb[, 1, drop = FALSE], cl))])
    cen  <- rowsum(emb, cl) / pmax(n_by, 1)
    lab_df <- data.frame(
      UMAP_1 = cen[, 1],
      UMAP_2 = cen[, 2],
      label  = rownames(cen),
      check.names = FALSE
    )

    p <- Seurat::DimPlot(obj, reduction = "umap", group.by = "seurat_clusters", label = FALSE) +
      ggplot2::ggtitle("UMAP (Seurat clusters)")

    if (requireNamespace("ggrepel", quietly = TRUE)) {
      p + ggrepel::geom_text_repel(
        data = lab_df, mapping = ggplot2::aes(UMAP_1, UMAP_2, label = label),
        size = 3, seed = 1, box.padding = 0.3, point.padding = 0.3
      )
    } else {
      p + ggplot2::geom_text(
        data = lab_df, mapping = ggplot2::aes(UMAP_1, UMAP_2, label = label),
        size = 3
      )
    }
  } else {
    # Absolute last resort: unlabeled UMAP
    Seurat::DimPlot(obj, reduction = "umap", group.by = "seurat_clusters", label = FALSE) +
      ggplot2::ggtitle("UMAP (Seurat clusters)")
  }
}


```


### UMAP (colored by annotation)

```{r umap-annot, fig.width=7, fig.height=6}
annot_png <- file.path(save_dir, "qc_umap_annotation.png")
if (file.exists(annot_png)) knitr::include_graphics(annot_png) else {
  plot.new(); text(0.5, 0.5, "Annotation UMAP not found.")
}
```

### Doublets

```{r doublets, fig.width=7, fig.height=6}
dbl_umap <- file.path(save_dir, "qc_umap_doublet.png")
dbl_bar1 <- file.path(save_dir, "qc_doublet_rate_by_group.png")

if (file.exists(dbl_umap)) knitr::include_graphics(dbl_umap) else {
  plot.new(); text(0.5, 0.5, "Doublet UMAP not found.")
}
if (file.exists(dbl_bar1)) knitr::include_graphics(dbl_bar1)

md <- obj@meta.data
if ("dbl_score" %in% colnames(md)) {
  show <- md[, c("dbl_score","is_doublet"), drop = FALSE]
  show$cell <- rownames(md)
  show <- head(show[order(show$dbl_score, decreasing = TRUE), ], 2000)
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(show, options = list(pageLength = 25, scrollX = TRUE))
  } else {
    knitr::kable(show, format = "html")
  }
} else {
  cat("> No doublet score available (pANN/scDblFinder.score).\n")
}
```

### Annotation validation (markers & GO)

```{r annot-validation}
heat_png <- file.path(save_dir, "celltype_gene_heatmap.png")
go_png   <- file.path(save_dir, "celltype_go_barplot.png")

if (file.exists(heat_png)) knitr::include_graphics(heat_png) else
  cat("> Marker heatmap not found (validation_plots_post_annotation not run or ComplexHeatmap not installed).\n")

if (file.exists(go_png)) knitr::include_graphics(go_png) else
  cat("> GO barplot not found (clusterProfiler/OrgDb not installed or no enriched terms).\n")

go2_png <- file.path(save_dir, "cluster_go_barplot.png")
if (file.exists(go2_png)) knitr::include_graphics(go2_png)
```

### Annotation confidence 

```{r annot-conf}
# choose an object variable provided to the template
obj_for_plot <- NULL
for (nm in c("obj","obj_keep","seu_qc","seu","obj_filtered","obj_all")) {
  if (exists(nm, inherits = FALSE)) { obj_for_plot <- get(nm); break }
}
if (!exists("imgDir", inherits = FALSE)) imgDir <- "qc_outputs"

if (!is.null(obj_for_plot)) {
  conf_png <- try(scQCenrich::scq_plot_auto_confidence(
    obj    = obj_for_plot,
    imgDir = imgDir
  ), silent = TRUE)

  if (!inherits(conf_png, "try-error") && is.character(conf_png) && file.exists(conf_png)) {
    knitr::include_graphics(conf_png)
  } else {
    cat("*(No numeric auto-annotation confidence found; skipping plot.)*")
  }
} else {
  cat("*(No Seurat object available for plotting.)*")
}

```

### QC Feature maps

```{r qc-featuremaps, fig.width=7, fig.height=6}
grid_png <- file.path(save_dir, "qc_featuremaps_grid.png")
indiv    <- Sys.glob(file.path(save_dir, "qc_featuremap_*.png"))

if (file.exists(grid_png)) {
  knitr::include_graphics(grid_png)
} else if (length(indiv)) {
  for (p in head(indiv, 12)) knitr::include_graphics(p)
} else {
  plot.new(); text(0.5, 0.5, "No featuremaps found. (qc_featuremaps() not run?)")
}
```

### Per-sample summary

```{r sample-summary, results="asis"}
suppressPackageStartupMessages({library(dplyr)})

df <- as.data.frame(metrics); df$cell <- rownames(df)
md <- obj@meta.data; md$cell <- rownames(md)
SCOL <- get0("SCOL", ifnotfound = NULL)

cols <- c("cell", "qc_status", "is_doublet", SCOL)
cols <- intersect(cols, colnames(md))
df <- dplyr::left_join(df, md[, cols, drop = FALSE], by = "cell")

if (!("qc_status" %in% colnames(df))) {
  sdf <- data.frame(cell = rownames(status_df), qc_status = status_df$qc_status, check.names = FALSE)
  df <- dplyr::left_join(df, sdf, by = "cell")
}

if (!("is_doublet" %in% colnames(df))) df$is_doublet <- NA

.pick_col <- function(df, md, candidates) {
  for (nm in candidates) {
    if (nm %in% colnames(df)) return(df[[nm]])
    if (nm %in% colnames(md)) return(md[[nm]])
  }
  NULL
}

assay <- Seurat::DefaultAssay(obj)

tc <- .pick_col(df, md, c("total_counts","libsize","lib_size","nCount_RNA","nCount","UMIs"))
if (is.null(tc)) { m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts"); tc <- setNames(as.numeric(Matrix::colSums(m)), colnames(m)) }
df$total_counts <- as.numeric(tc[df$cell])

ng <- .pick_col(df, md, c("n_features","nGenes","nFeature_RNA","nFeature"))
if (is.null(ng)) { m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts"); ng <- setNames(as.numeric(Matrix::colSums(m > 0)), colnames(m)) }
df$n_features <- as.numeric(ng[df$cell])

pm <- .pick_col(df, md, c("pct_mito","percent.mt","percent_mito","mito_percent"))
if (is.null(pm)) {
  m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts")
  idx <- .mt_gene_indices(rownames(m), species = get0("species", ifnotfound = "mouse"))
  pm <- if (length(idx)) 100 * Matrix::colSums(m[idx,,drop=FALSE]) / pmax(Matrix::colSums(m),1)
        else setNames(rep(NA_real_, ncol(m)), colnames(m))
}
df$pct_mito <- as.numeric(pm[df$cell])

grp <- if (!is.null(SCOL) && SCOL %in% colnames(df)) dplyr::group_by(df, .data[[SCOL]]) else dplyr::group_by(df, grp = "All cells")

sum_tbl <- grp %>%
  dplyr::summarise(
    cells_total  = dplyr::n(),
    cells_keep   = sum(qc_status == "keep", na.rm=TRUE),
    cells_remove = sum(qc_status == "remove", na.rm=TRUE),
    libsize_med  = median(total_counts, na.rm=TRUE),
    genes_med    = median(n_features,   na.rm=TRUE),
    mito_med     = median(pct_mito,     na.rm=TRUE),
    dbl_rate     = mean(as.logical(is_doublet), na.rm = TRUE)
  ) %>% dplyr::ungroup()

if (nrow(sum_tbl) == 0) {
  cat("> No cells to summarise.\n")
} else if (requireNamespace("gt", quietly = TRUE)) {
  tab <- gt::gt(sum_tbl) |>
    gt::fmt_number(columns = c(libsize_med, genes_med, mito_med), decimals = 1) |>
    gt::cols_label(
      cells_total  = "Cells", cells_keep = "Kept", cells_remove = "Removed",
      libsize_med  = "Median counts", genes_med = "Median genes",
      mito_med     = "Median %mito", dbl_rate = "Doublet rate"
    ) |>
    gt::tab_header(title = gt::md("QC Summary")) |>
    gt::opt_table_font(stack = "system")
  cat(gt::as_raw_html(tab))
} else if (requireNamespace("DT", quietly = TRUE)) {
  print(DT::datatable(sum_tbl, rownames = FALSE, options = list(pageLength = 10)))
} else {
  cat(knitr::kable(sum_tbl, format = "html"))
}
```

### Distributions by QC status

```{r dists, fig.width=7, fig.height=6}
library(ggplot2)
cells <- colnames(obj)
assay <- Seurat::DefaultAssay(obj)

get_total_counts <- function() {
  if (!is.null(metrics) && "nCount" %in% names(metrics)) {
    v <- setNames(as.numeric(metrics$nCount), rownames(metrics))
    return(v[cells])
  }
  if ("nCount_QC" %in% colnames(obj@meta.data)) {
    v <- setNames(as.numeric(obj@meta.data$nCount_QC), rownames(obj@meta.data))
    return(v[cells])
  }
  m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts")
  v <- setNames(as.numeric(Matrix::colSums(m)), colnames(m))
  v[cells]
}

get_n_features <- function() {
  if (!is.null(metrics) && "nFeature" %in% names(metrics)) {
    v <- setNames(as.numeric(metrics$nFeature), rownames(metrics))
    return(v[cells])
  }
  if ("nFeature_QC" %in% colnames(obj@meta.data)) {
    v <- setNames(as.numeric(obj@meta.data$nFeature_QC), rownames(obj@meta.data))
    return(v[cells])
  }
  m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts")
  v <- setNames(as.numeric(Matrix::colSums(m > 0)), colnames(m))
  v[cells]
}

get_pct_mito <- function() {
  if (!is.null(metrics) && "pctMT" %in% names(metrics)) {
    v <- setNames(as.numeric(100 * metrics$pctMT), rownames(metrics))
    return(v[cells])
  }
  m <- Seurat::GetAssayData(obj, assay = assay, layer = "counts")
  idx <- grep("^mt-", rownames(m), ignore.case = TRUE)
  if (!length(idx)) {
    # optional: try your internal helper if the package is loaded
    idx <- tryCatch(scQCenrich:::.mt_gene_indices(rownames(m),
              species = get0("species","mouse")), error = function(...) integer(0))
  }
  out <- if (length(idx)) 100 * Matrix::colSums(m[idx,,drop=FALSE]) /
           pmax(Matrix::colSums(m), 1) else rep(NA_real_, ncol(m))
  setNames(as.numeric(out), colnames(m))[cells]
}


total_counts <- get_total_counts()
n_features   <- get_n_features()
pct_mito     <- get_pct_mito()
qc_status    <- obj@meta.data[cells, "qc_status", drop = TRUE]

long <- data.frame(
  total_counts = total_counts,
  n_features   = n_features,
  pct_mito     = pct_mito,
  qc_status    = qc_status,
  row.names    = cells, check.names  = FALSE
)
long <- long[complete.cases(long[c("total_counts","n_features","pct_mito")]) & !is.na(long$qc_status), , drop = FALSE]

p1 <- ggplot(long, aes(total_counts, fill = qc_status)) + geom_histogram(bins = 60, alpha = .8) + theme_classic() + labs(title = "Library size")
p2 <- ggplot(long, aes(n_features,   fill = qc_status)) + geom_histogram(bins = 60, alpha = .8) + theme_classic() + labs(title = "#Features")
p3 <- ggplot(long, aes(pct_mito,     fill = qc_status)) + geom_histogram(bins = 60, alpha = .8) + theme_classic() + labs(title = "% Mito")

print(p1); print(p2); print(p3)
```

### Removed cells (top 2,000 rows)

```{r removed}
removed_cells <- subset(obj@meta.data, qc_status == "remove")
removed_cells$cell <- rownames(removed_cells)
show_tbl <- head(removed_cells, 2000)
if (nrow(show_tbl) == 0) {
  cat("> No cells were removed.\n")
} else if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(show_tbl, options = list(pageLength = 25, scrollX = TRUE))
} else {
  knitr::kable(show_tbl, format = "html")
}
```

---

#### Session info

```{r}
sessionInfo()
```

```

---
